Jovian enables direct inference of germline haplotypes from short reads via sequence-to-sequence modeling

### 2022-10-14 manuscript/manuscript.tex
No vaftgt or parallel decomp

\title{Jovian enables direct inference of germline haplotypes from short reads via sequence-to-sequence modeling}
\author[1]{Brendan O'Fallon}
\author[1]{Ashini Bolia}
\author[1]{Jacob Durtschi}
\author[1]{Luobin Yang}
\author[1]{Eric Frederickson}
\author[1]{Katherine Noble}
\author[1]{Joshua Coleman}
\author[1]{Hunter Best}

\affil[1]{ARUP Institute for Clinical and Experimental Pathology, Salt Lake City, UT}
\date{}
\begin{document}

@@

\begin{abstract}
	Detection of germline variants in next-generation sequencing data is an essential component of modern genomics analysis. Variant detection tools typically rely on statistical algorithms such as de Bruijn graphs, Hidden Markov Models and regression models, often coupled with heuristic techniques and thresholds to identify variants. Here we describe a new approach that replaces these handcrafted statistical methods with a single, end-to-end deep learning model that directly infers germline haplotypes from short read pileups. Our model, called Jovian, frames variant detection as a sequence-to-sequence modeling task, akin to language translation, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences. After training with 17 whole genome sequences from Genome-in-a-Bottle cell lines, we demonstrate that this method learns to realign complex and ambiguous read mappings to produce accurate haplotype predictions, predicts variant genotypes and phase accurately, and leverages the local read context to inform predictions about a given position. We also demonstrate that a 2-dimensional positional encoding significantly improved precision of the detected variants. Compared to other callers, sensitivity and precision is higher than GATK HaplotypeCaller, but lower than DeepVariant and Strelka2. 
\end{abstract}

@@

\begin{center}
	\begin{table}
	\begin{tabular}{ l }
	 \textbf{Feature} \\
	 Mean probability of base call at all positions affected by variant \\ 
	 Number of reference bases affected \\
	 Number of alternate bases affected \\
	 Minimum probability of base call at all positions affected by variant \\
	 Maximum probability of base call at all positions affected by variant \\
	 Number of variants in \textit{cis} (same haplotype) with this variant \\
	 Number of variants in \textit{trans} (other haplotype) with this variant \\ 
	 Total number of variants in both \textit{cis} and \textit{trans} \\
	 Distance in bases of call from beginning of window - minimum across windows \\
	 Distance in bases of call from beginning of window - maximum across windows \\
	 Read depth at variant position \\
	 Gnomad genome frequency (v2.0.2) of variant \\
	 Zygosity (heterozygous or homozygous) of variant \\
	 Fraction of reads in BAM/CRAM supporting variant call \\
	 Fraction of reads with mapping quality > 10 supporting variant call \\
	 Total number of reads supporting variant call \\
	\end{tabular}
	\label{table:features}
	\caption{ Features used to train classifier model for estimation of variant quality scores }
\end{table}
\end{center}

@@

Our model uses an unmodified transformer encoder with a very simple decoder consisting of just two fully connected layers, and we leave to future work exploration of the many transformer variations proposed in recent years (e.g. Dosovitskiy 2018, Fedus et al. 2021, Liu et al. 2021, Wu et al. 2021). Our simple decoder has a fixed length equal to the number of input tokens (genomic positions), but a natural improvement would be to add a more sophisticated decoder capable of producing a variable length sequence to handle larger insertion and deletion events.  In addition, newer transformer models that avoid the $O(n^2)$ scaling with the number of input tokens may be able to handle longer input sequence lengths, and therefore may be able to detect larger variants.

>>>>

\title{Jovian enables direct inference of germline haplotypes from short reads via sequence-to-sequence modeling}
\author{Author names witheld during review}
%\author[1]{Brendan O'Fallon}
%\author[1]{Ashini Bolia}
%\author[1]{Jacob Durtschi}
%\author[1]{Luobin Yang}
%\author[1]{Eric Frederickson}
%\author[1]{Katherine Noble}
%\author[1]{Joshua Coleman}
%\author[1]{Hunter Best}

%\affil[1]{ARUP Institute for Clinical and Experimental Pathology, Salt Lake City, UT}
\date{}
\begin{document}

@@

\begin{abstract}
	Detection of germline variants in next-generation sequencing data is an essential component of modern genomics analysis. Variant detection tools typically rely on statistical algorithms such as de Bruijn graphs, Hidden Markov Models and regression models, often coupled with heuristic techniques and thresholds to identify variants. Here we describe a new approach that replaces these handcrafted statistical methods with a single, end-to-end deep learning model that directly infers germline haplotypes from short read pileups. Our model, called Jovian, frames variant detection as a sequence-to-sequence modeling task, akin to language translation, and employs a transformer-based architecture to translate alignment columns into two predicted haplotype sequences. After training with 17 whole genome sequences from Genome-in-a-Bottle cell lines, we demonstrate that this method learns to realign complex and ambiguous read mappings to produce accurate haplotype predictions, infers variant genotype and phase accurately, and leverages the local read context to inform predictions about a given position. We also demonstrate that a 2-dimensional positional encoding significantly improved precision of the detected variants. Compared to other callers, sensitivity and precision are higher than GATK HaplotypeCaller, but lower than DeepVariant and Strelka2. 
\end{abstract}

@@

\begin{center}
	\begin{table}
	\begin{tabular}{ l }
	 \textbf{Feature} \\
	 Mean probability of base call at all positions affected by variant \\ 
	 Number of reference bases affected \\
	 Number of alternate bases affected \\
	 Minimum probability of base call at all positions affected by variant \\
	 Maximum probability of base call at all positions affected by variant \\
	 Number of variants in \textit{cis} (same haplotype) with this variant \\
	 Number of variants in \textit{trans} (other haplotype) with this variant \\ 
	 Total number of variants in both \textit{cis} and \textit{trans} \\
	 Distance in bases of call from beginning of window - minimum across windows \\
	 Distance in bases of call from beginning of window - maximum across windows \\
	 Read depth at variant position \\
	 Gnomad genome frequency (v2.0.2) of variant \\
	 Zygosity (heterozygous or homozygous) of variant \\
	 Fraction of reads in BAM/CRAM supporting variant call \\
	 Fraction of reads with mapping quality $>$ 10 supporting variant call \\
	 Total number of reads supporting variant call \\
	\end{tabular}
	\label{table:features}
	\caption{ Features used to train classifier model for estimation of variant quality scores }
\end{table}
\end{center}

@@

Our model uses an unmodified transformer encoder with a very simple decoder consisting of just two fully connected layers, and we leave to future work exploration of the many transformer variations proposed in recent years (e.g. Dosovitskiy 2018, Fedus et al. 2021, Liu et al. 2021, Wu et al. 2021). Our simple decoder has a fixed length equal to the number of input tokens (genomic positions), but a natural improvement would be to add a more sophisticated autoregressive decoder capable of producing a variable length sequence to handle larger insertion and deletion events.  In addition, newer transformer models that avoid the $O(n^2)$ scaling with the number of input tokens may be able to handle longer input sequence lengths, and therefore may be able to detect larger variants.
